Appendix: Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2025-01-13 appendix.qmd
Added the notations
Added the equations with a flow
Updated the appendix
Moved scDEED to appendix

---
title: "Appendix: Looking at Non-Linear Dimension Reductions as Models in the Data Space"
format: 
    jasa-pdf:
        keep-tex: true
    jasa-html: default
author:
  - name: Jayani P.G. Lakshika
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC 
        country: Australia
        postal-code: 3800
    orcid: 0000-0002-6265-6481
    email: jayani.piyadigamage@monash.edu
    url: https://jayanilakshika.netlify.app/
  - name: Dianne Cook
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC 
        country: Australia
        postal-code: 3800
    orcid: 0000-0002-3813-7155
    email: dicook@monash.edu 
    url: http://www.dicook.org/
  - name: Paul Harrison
    affiliations:
      - name: Monash University
        department: MGBP, BDInstitute
        address: Clayton
        city: VIC 
        country: Australia
        postal-code: 3800
    orcid: 0000-0002-3980-268X
    email: 	paul.harrison@monash.edu
    url: 
  - name: Michael Lydeamore
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC 
        country: Australia
        postal-code: 3800
    orcid: 0000-0001-6515-827X
    email: michael.lydeamore@monash.edu
    url: https://www.michaellydeamore.com/
  - name: Thiyanga S. Talagala
    affiliations:
      - name: University of Sri Jayewardenepura
        department: Statistics
        address: Gangodawila
        city: Nugegoda 
        country: Sri Lanka
        postal-code: 10100
    orcid: 0000-0002-0656-9789
    email: ttalagala@sjp.ac.lk 
    url: https://thiyanga.netlify.app/
tbl-cap-location: bottom
header-includes: | 
  \usepackage{amsmath}
  \usepackage{float}
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \usepackage{bm}
  \def\tightlist{}
  \usepackage{setspace}
  \newcommand\pD{$p\text{-}D$}
  \newcommand\kD{$k\text{-}D$}
  \newcommand\dD{$d\text{-}D$}
  \newcommand\gD{$2\text{-}D$}
---

@@

```{r}

triangle_temp <- tibble(
  x = c(0, 4, 0),
  y = c(0, 0, 3)
)

@@

```{r}
#| out-height: 30%
triangle_plot
```

By pythagorean theorem,

$$
l^2 = (\frac{a_1}{2})^2 + (\frac{a_2 - l}{2})^2
$$
$$
l^2 - (\frac{a_2 - l}{2})^2 = (\frac{a_1}{2})^2
$$
$$
[l - (\frac{a_2 - l}{2})][l + (\frac{a_2 - l}{2})] = (\frac{a_1}{2})^2
$$

$$
3l^2 + 2la_2 - (a_1^2 + a_2^2) = 0
$$

$$
l = \frac{-2a_2 \pm \sqrt{4a_2^2 - 24[-(a_1^2 + a_2^2)]}}{6}
$$
$$
l = \frac{-a_2 \pm \sqrt{a_2^2 - 6[-(a_1^2 + a_2^2)]}}{3}
$$

>>>>

---
title: "Appendix: Looking at Non-Linear Dimension Reductions as Models in the Data Space"
format: 
    jasa-pdf:
        keep-tex: true
    jasa-html: default
author:
  - name: Jayani P.G. Lakshika
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC 
        country: Australia
        postal-code: 3800
    orcid: 0000-0002-6265-6481
    email: jayani.piyadigamage@monash.edu
    url: https://jayanilakshika.netlify.app/
  - name: Dianne Cook
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC 
        country: Australia
        postal-code: 3800
    orcid: 0000-0002-3813-7155
    email: dicook@monash.edu 
    url: http://www.dicook.org/
  - name: Paul Harrison
    affiliations:
      - name: Monash University
        department: MGBP, BDInstitute
        address: Clayton
        city: VIC 
        country: Australia
        postal-code: 3800
    orcid: 0000-0002-3980-268X
    email: 	paul.harrison@monash.edu
    url: 
  - name: Michael Lydeamore
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC 
        country: Australia
        postal-code: 3800
    orcid: 0000-0001-6515-827X
    email: michael.lydeamore@monash.edu
    url: https://www.michaellydeamore.com/
  - name: Thiyanga S. Talagala
    affiliations:
      - name: University of Sri Jayewardenepura
        department: Statistics
        address: Gangodawila
        city: Nugegoda 
        country: Sri Lanka
        postal-code: 10100
    orcid: 0000-0002-0656-9789
    email: ttalagala@sjp.ac.lk 
    url: https://thiyanga.netlify.app/
tbl-cap-location: bottom
header-includes: | 
  \usepackage{amsmath}
  \usepackage{float}
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \usepackage{bm}
  \def\tightlist{}
  \usepackage{setspace}
  \newcommand\pD{$p\text{-}D$}
  \newcommand\kD{$k\text{-}D$}
  \newcommand\dD{$d\text{-}D$}
  \newcommand\gD{$2\text{-}D$}
  \newcommand{\To}{\textbf{to}~}
  \usepackage{abstract}
  \renewcommand{\abstractname}{}    % clear the title
  \renewcommand{\absnamepos}{empty} % originally center
---

@@

```{r}
# triangle_temp <- tibble(
#   x = c(0, 4, 0),
#   y = c(0, 0, 3)
# )

@@

```{r}
#| label: fig-tri-param
#| fig-cap: "The components of the right triangle illustrating notation."
#| out-height: 30%

triangle_plot
```

By applying the Pythagorean theorem, we obtain,

$$
l^2 = \left(\frac{a_1}{2}\right)^2 + \left(\frac{a_2 - l}{2}\right)^2.
$$
Next, rearranging the terms, we get,

$$
l^2 - \left(\frac{a_2 - l}{2}\right)^2 = \left(\frac{a_1}{2}\right)^2,
$$

$$
\left[l - \left(\frac{a_2 - l}{2}\right)\right]\left[l + \left(\frac{a_2 - l}{2}\right)\right] = \left(\frac{a_1}{2}\right)^2,
$$

$$
3l^2 + 2a_2l - (a_1^2 + a_2^2) = 0.
$$

Finally, by solving the quadratic equation, we compute,

$$
l = \frac{-2a_2 \pm \sqrt{4a_2^2 - 24[-(a_1^2 + a_2^2)]}}{6},
$$

$$
l = \frac{-a_2 \pm \sqrt{a_2^2 - 6[-(a_1^2 + a_2^2)]}}{3},
$$

where $l > 0$.

#### Comparison with results of scDEED recommendations

<!--
XXX What this section needs:

- connect this with previous section, not make it seem like a completely new. connections are: nldr used to show clustering, scDEED provides a different approach to decide which is best.
- why these two? assume one is the published figure, and second is one recommended by scDEED, which our results confirm is better.
- Refer to figure in text
- Fig 16 not necessary

CHANGES WERE MADE BASED ON THIS THINKING
-->

<!-- In the field of single-cell studies, clustering is a common analytical task used to identify groups of cells with similar expression profiles. Non-linear dimensional reduction (NLDR) methods are frequently employed to visualize these clusters and help validate the results. However, it is well known that the 2D embeddings produced by t-SNE and UMAP may not accurately reflect the similarities among cell clusters. ALREADY STATED IN PREVIOUS SECTION -->

As we were writing this paper @xia2023 appeared proposing a new method called scDEED helping to assess the validity of a \gD{} embedding. scDEED calculates a reliability score for each cell embedding based on the similarity between the cell’s \gD{} embedding neighbors and its neighbors prior to embedding. A low reliability score suggests a dubious embedding. It can help in the deciding on optimal hyperparameters. Here we illustrate how our method compares with the results from scDEED

Following the process in @xia2023 again using the Human Peripheral Blood Mononuclear Cells (PBMC) data. Note that @xia2023 uses a different subset from @sec-pbmc, which contains $31,021$ cells including cell type labels, and the gene expression levels were in the unit of log-transformed UMI count per $10,000$. They focused on three sequencing methods (inDrops, DropSeq, and SeqWell) and four common cell types Cytotoxic T cell, CD4+T cell, CD14+ Monocyte, and B cell.

For illustration purposes, we only selected cells generated with inDrops ($n=5858$ cells) and UMAP cell embeddings. Also, @xia2023 used first $50$ principal components to generate the UMAP. The objective is to assess the optimized layout by scDEED, and if it does not accurately represent the three clusters with small separations of the PBMC dataset, then select a reasonable \gD{} layout.

@fig-model-pbmc-author-proj-scdeed shows layout from the original published article along with that suggested by scDEED to be more accurate. The log MSE vs binwidth plot illustrates that our approach would suggest that scDEED is correct here, that layout b is more accurately reflecting the cluster structure in the data.

```{r}
result_umap <- read_rds(here::here('data/pbmc3k/pbmc_scdeed_umap_results.rds'))
```

```{r}
#| eval: false
result_umap |>
  kableExtra::kable(format = "latex",
                    booktabs = TRUE,
                    longtable = TRUE,
                    label = "scdeedumap") |>
  kableExtra::kable_styling(latex_options = "scale_down")
```

```{r}
#| out-width: 50%
#| eval: false

result_umap |>
  filter(min.dist == 0.5) |>
  ggplot(aes(
           x = as.numeric(n_neighbors),
           y = number_dubious_cells
         )) +
    geom_line() +
    geom_point() +
    labs(x = "n_neighbors",
         y = "# of dubious cells") +
    theme_minimal()
```

```{r}
#| label: read-pbmc-umap-layouts
# Read a variety of different NLDR representations of PBMC
# and plot them on same aspect ratio
umap_pbmc1 <- read_rds("data/pbmc3k/pbmc_scdeed_umap_n_neighbors_30_min_dist_0.3.rds")
umap_pbmc1 <- as_tibble(umap_pbmc1)
names(umap_pbmc1) <- c("UMAP1", "UMAP2")

nldr1c <- umap_pbmc1 |>
  ggplot(aes(x = UMAP1,
             y = UMAP2)) +
  geom_point(alpha=0.1, size=1, colour='#e41a1c') +
  interior_annotation("a")

umap_pbmc2 <- read_rds("data/pbmc3k/pbmc_scdeed_umap_n_neighbors_80_min_dist_0.5.rds")
umap_pbmc2 <- as_tibble(umap_pbmc2)
names(umap_pbmc2) <- c("UMAP1", "UMAP2")

nldr2c <- umap_pbmc2 |>
  ggplot(aes(x = UMAP1,
             y = UMAP2))+
  geom_point(alpha=0.1, size=1, colour='#377eb8') +
  interior_annotation("b")

```{r}
#| label: combine-data-pbmc-umap

error_pbmc_umap <- read_rds("data/pbmc3k/error_scdeed_pbmc_umap_30_min_dist_0.3.rds")
error_pbmc_umap2 <- read_rds("data/pbmc3k/error_scdeed_pbmc_umap_80_min_dist_0.5.rds")

error_pbmc <- bind_rows(error_pbmc_umap, 
                        error_pbmc_umap2)

error_pbmc <- error_pbmc |>
  mutate(a1 = round(a1, 2)) |>
  filter(bin1 >= 5) |>
  group_by(method, a1) |>
  filter(MSE == min(MSE)) |>
  ungroup()
```

```{r}
#| label: error-comp-pbmc-umap

error_umap_plot_pbmc <- ggplot(error_pbmc, 
                               aes(x = a1, 
                                   y = MSE, 
                                   colour = method)) + 
  geom_point(size = 0.8) +
  geom_line(linewidth = 0.3) + 
  # geom_vline(xintercept = 15, linetype="solid", 
  #            color = "black", linewidth=0.8, alpha = 0.5) +
  scale_x_continuous(breaks = sort(unique(error_pbmc$a1))[c(1, 5, 9, 12, 15, 17, 19)]) +
  scale_color_manual(values=c('#e41a1c','#377eb8')) +
  scale_y_log10() +
  ylab("log(MSE)") +
  xlab(expression(paste("binwidth (", a[1], ")"))) +
  theme_minimal() +
  theme(panel.border = element_rect(fill = 'transparent'),
        plot.title = element_text(size = 12, hjust = 0.5, vjust = -0.5),
        axis.ticks.x = element_line(),
        axis.ticks.y = element_line(),
        legend.position = "none",
        axis.text.x = element_text(size = 7),
        axis.text.y = element_text(size = 7),
        axis.title.x = element_text(size = 7),
        axis.title.y = element_text(size = 7))

```{r}
#| label: fig-pbmc-mse-umap
#| fig-pos: H
#| out-width: 100%
#| fig-cap: "Assessing which of the 2 UMAP layouts with different hyperparameter settings (n_neighbors: $30$, min_dist: $0.3$ (red), n_neighbors: $80$, min_dist: $0.5$ (blue)) on the PBMC data is the better representation using MSE for varying binwidth ($a_1$). Colour used for the lines and points in the left plot and in the scatterplots represents UMAP layout (a, b). Of the two, layout b is optimal across all binwidths making it the best choice."

free(error_umap_plot_pbmc) + wrap_plots(nldr1c, nldr2c, ncol = 2) +
  plot_layout(ncol = 1, heights = c(1, 1))
```

```{r}
#| label: published-pbmc-scdeed

training_data_pbmc <- read_rds("data/pbmc3k/pbmc_pca_50_scdeed.rds")
training_data_pbmc <- training_data_pbmc[, 1:50] |>
  mutate(ID = 1:NROW(training_data_pbmc))

## For umap
umap_pbmc <- read_rds("data/pbmc3k/pbmc_scdeed_umap_n_neighbors_80_min_dist_0.5.rds")
umap_pbmc <- as_tibble(umap_pbmc)
names(umap_pbmc) <- c("UMAP1", "UMAP2")
umap_pbmc <- umap_pbmc |>
  mutate(ID = 1:NROW(umap_pbmc))

pbmc_scaled_obj_umap <- gen_scaled_data(
  data = umap_pbmc)
umap_pbmc_scaled <- pbmc_scaled_obj_umap$scaled_nldr

lim1 <- pbmc_scaled_obj_umap$lim1
lim2 <- pbmc_scaled_obj_umap$lim2
r2_umap <- diff(lim2)/diff(lim1)

```{r}
#| label: hexbin-pbmc-scdeed
## Compute hexbin parameters
num_bins_x_pbmc <- 12
lim1 <- pbmc_scaled_obj_umap$lim1
lim2 <- pbmc_scaled_obj_umap$lim2
r2_pbmc <- diff(lim2)/diff(lim1)

pbmc_model <- fit_highd_model(
  training_data = training_data_pbmc,
  emb_df = umap_pbmc_scaled,
  bin1 = num_bins_x_pbmc,
  r2 = r2_pbmc,
  q = 0.1,
  is_bin_centroid = TRUE,
  is_rm_lwd_hex = FALSE,
  col_start_highd = "pc"
)

df_bin_centroids_pbmc <- pbmc_model$df_bin_centroids
df_bin_pbmc <- pbmc_model$df_bin

## Triangulate bin centroids
tr1_object_pbmc <- tri_bin_centroids(
  df_bin_centroids_pbmc, x = "c_x", y = "c_y")
tr_from_to_df_pbmc <- gen_edges(
  tri_object = tr1_object_pbmc)

## Compute 2D distances
distance_pbmc <- cal_2d_dist(
  tr_coord_df = tr_from_to_df_pbmc,
  start_x = "x_from",
  start_y = "y_from",
  end_x = "x_to",
  end_y = "y_to",
  select_vars = c("from", "to", "distance"))

## To find the benchmark value
benchmark_pbmc <- find_lg_benchmark(
  distance_edges = distance_pbmc,
  distance_col = "distance")

tr_df <- distinct(tibble(
  x = c(tr_from_to_df_pbmc[["x_from"]], tr_from_to_df_pbmc[["x_to"]]), 
  y = c(tr_from_to_df_pbmc[["y_from"]], tr_from_to_df_pbmc[["y_to"]])))

distance_df_small_edges_pbmc <- distance_pbmc |>
  filter(distance < benchmark_pbmc)

tr_from_to_df_pbmc <- inner_join(
  tr_from_to_df_pbmc, distance_df_small_edges_pbmc, 
  by = c("from", "to"))

trimesh_removed_pbmc <- ggplot() + 
  geom_segment(data = tr_from_to_df_pbmc, 
               aes(
                 x = x_from, 
                 y = y_from, 
                 xend = x_to, 
                 yend = y_to),
               colour = "#000000",
               linewidth = 1) +
  geom_point(
    data = umap_pbmc_scaled,
    aes(
      x = UMAP1,
      y = UMAP2
    ),
    alpha = 0.2,
    color = '#a65628'
  )  +
  interior_annotation("a1", cex = 2) +
  theme(
    aspect.ratio = 1
  )
```

```{r}
#| label: pbmc-umap-model-proj-scdeed

data_pbmc <- training_data_pbmc |> 
  select(-ID) |>
  mutate(type = "data")

df_b_pbmc <- df_bin_pbmc |>
  dplyr::filter(hb_id %in% df_bin_centroids_pbmc$hexID) |>
  dplyr::mutate(type = "model") ## Data with summarized mean

## Reorder the rows of df_b according to the hexID order in df_b_with_center_data
df_b_pbmc <- df_b_pbmc[match(df_bin_centroids_pbmc$hexID, df_b_pbmc$hb_id),] |>
  dplyr::select(-hb_id)

# Apply the scaling
df_model_data_pbmc <- bind_rows(data_pbmc, df_b_pbmc)
scaled_pbmc <- scale_data_manual(df_model_data_pbmc, "type") |>
  as_tibble()

scaled_pbmc_data <- scaled_pbmc |>
  filter(type == "data") |>
  select(-type)

scaled_pbmc_data_model <- scaled_pbmc |>
  filter(type == "model") |>
  select(-type)

# Combine with the true model for visualization
df <- dplyr::bind_rows(scaled_pbmc_data_model |> mutate(type = "model"),
                       scaled_pbmc_data |> mutate(type = "data"))

## First projection
projection <- cbind(
  c(0.040856,0.000723,0.006000,0.007623,0.001986,-0.006599,0.005311,-0.000222,0.001648,0.008067,-0.001587,0.004423,0.004851,0.000638,0.002606,0.006616,0.006188,-0.000954,-0.004603,0.004998,-0.012220,0.008401,-0.007788,0.008972,-0.003903,0.003348,-0.001817,-0.000385,-0.002437,-0.004097,-0.001357,0.013660,0.001770,0.006982,0.002828,-0.009169,-0.003541,0.008582,-0.000032,-0.002375,0.005854,-0.005466,-0.000824,-0.001435,0.004148,0.007141,0.000883,-0.004159,0.005401,0.007882),
  c(-0.000590,0.039365,-0.004481,-0.009041,-0.003805,0.003278,-0.009601,-0.000650,-0.002333,-0.005697,-0.001488,-0.009808,0.002557,0.004113,0.005705,-0.004665,-0.001170,-0.004727,-0.000845,0.005084,-0.018054,0.002538,0.006606,-0.007184,0.000238,0.001877,-0.002124,-0.005972,-0.009729,0.004958,-0.003181,-0.005185,-0.009598,-0.004243,0.007716,0.000396,-0.004258,0.005932,-0.003242,-0.004184,0.003493,-0.008405,0.000860,0.008304,0.004569,0.000798,0.004156,-0.005777,0.005278,-0.002531))

projection_scaled <- projection * 3

# projection_filtered <- projection
# projection_filtered[apply(projection_filtered, 1, function(row) all(row < 0)), ] <- 0.75

#rows_to_keep <- c(1, 2, 4, 5, 6)

#projection_filtered <- projection[rows_to_keep, ]

projected <- as.matrix(scaled_pbmc_data) %*% projection_scaled

projected_df <- projected |>
  tibble::as_tibble(.name_repair = "unique") |>
  dplyr::rename(c("proj1" = "...1",
                  "proj2" = "...2")) |>
  #dplyr::mutate(type = df_exe$type) |>
  dplyr::mutate(ID = dplyr::row_number())

projected_model <- as.matrix(scaled_pbmc_data_model) %*% projection_scaled

projected_model_df <- projected_model |>
  tibble::as_tibble(.name_repair = "unique") |>
  dplyr::rename(c("proj1" = "...1",
                  "proj2" = "...2")) |>
  dplyr::mutate(ID = dplyr::row_number())

model_df <- dplyr::left_join(
  distance_df_small_edges_pbmc |> select(-distance), 
  projected_model_df, 
  by = c("from" = "ID"))

names(model_df)[3:NCOL(model_df)] <- paste0(names(projected_model_df)[-NCOL(projected_model_df)], "_from")

model_df <- dplyr::left_join(model_df, projected_model_df, by = c("to" = "ID"))
names(model_df)[(2 + NCOL(projected_model_df)):NCOL(model_df)] <- paste0(names(projected_model_df)[-NCOL(projected_model_df)], "_to")

axes_obj <- gen_axes(
  proj = projection,
  limits = 0.04,
  axis_pos_x = -0.025,
  axis_pos_y = -0.025,
  axis_labels = names(scaled_pbmc_data),
  threshold = 0.0001)

axes <- axes_obj$axes
circle <- axes_obj$circle

pbmc_proj_umap_model1 <- projected_df |>
  ggplot(
    aes(
      x = proj1, 
      y = proj2)) +
  geom_segment(
    data = model_df, 
    aes(
      x = proj1_from, 
      y = proj2_from, 
      xend = proj1_to, 
      yend = proj2_to), 
    color = "#000000") +
  geom_point(
    #size = 0.5,
    alpha = 0.2,
    color = '#a65628') +
  geom_segment(
    data=axes, 
    aes(x=x1, y=y1, xend=x2, yend=y2), 
    colour="grey70") +
  geom_text(
    data=axes,
    aes(x=x2, y=y2), 
    label=rownames(axes),
    colour="grey50",
    size = 3) +
  geom_path(
    data=circle, 
    aes(x=c1, y=c2), colour="grey70") +
  coord_fixed() +
  xlim(c(-0.03, 0.06)) +
  ylim(c(-0.03, 0.06)) +
  interior_annotation("a2", cex = 2)

## Second projection
projection <- cbind(
  c(-0.019825,-0.011101,0.011543,-0.006943,-0.004485,0.000062,0.002023,-0.005178,0.000800,-0.002903,-0.014182,0.008756,0.009595,0.007032,-0.007820,0.006143,0.004865,0.003352,0.007554,0.002265,0.012035,0.009961,-0.001825,-0.004942,-0.002237,0.002308,0.009531,-0.001146,0.013401,-0.001022,0.008434,0.000477,0.004239,-0.006772,-0.004702,-0.011432,-0.000431,0.008146,-0.009299,-0.004782,0.003457,-0.008214,-0.009604,0.011329,0.003805,-0.003004,-0.000800,0.015288,-0.013157,0.006984),
  c(-0.001387,0.001643,0.010470,0.002930,-0.002335,-0.010569,0.013839,0.002917,-0.006825,0.010191,-0.006367,0.002220,0.002506,0.005139,-0.007213,-0.006235,-0.007123,-0.000222,-0.002225,0.011592,-0.003942,0.002788,-0.000055,0.011316,-0.009753,0.020785,0.004615,0.005172,0.011182,-0.003125,-0.010860,0.007820,-0.017510,0.007010,0.007267,0.002694,-0.003949,-0.009188,-0.005163,0.001898,-0.004744,0.013757,0.010630,-0.007688,0.004598,0.004658,-0.011154,0.003584,-0.005226,0.006650))

projection_scaled <- projection * 3

projected <- as.matrix(scaled_pbmc_data) %*% projection_scaled

projected_df <- projected |>
  tibble::as_tibble(.name_repair = "unique") |>
  dplyr::rename(c("proj1" = "...1",
                  "proj2" = "...2")) |>
  #dplyr::mutate(type = df_exe$type) |>
  dplyr::mutate(ID = dplyr::row_number())

projected_model <- as.matrix(scaled_pbmc_data_model) %*% projection_scaled

projected_model_df <- projected_model |>
  tibble::as_tibble(.name_repair = "unique") |>
  dplyr::rename(c("proj1" = "...1",
                  "proj2" = "...2")) |>
  dplyr::mutate(ID = dplyr::row_number())

model_df <- dplyr::left_join(
  distance_df_small_edges_pbmc |> select(-distance), 
  projected_model_df, 
  by = c("from" = "ID"))

names(model_df)[3:NCOL(model_df)] <- paste0(names(projected_model_df)[-NCOL(projected_model_df)], "_from")

model_df <- dplyr::left_join(model_df, projected_model_df, by = c("to" = "ID"))
names(model_df)[(2 + NCOL(projected_model_df)):NCOL(model_df)] <- paste0(names(projected_model_df)[-NCOL(projected_model_df)], "_to")

axes_obj <- gen_axes(
  proj = projection,
  limits = 0.03,
  axis_pos_x = -0.02,
  axis_pos_y = -0.02,
  axis_labels = names(scaled_pbmc_data),
  threshold = 0)

axes <- axes_obj$axes
circle <- axes_obj$circle

pbmc_proj_umap_model2 <- projected_df |>
  ggplot(
    aes(
      x = proj1,
      y = proj2)) +
  geom_segment(
    data = model_df,
    aes(
      x = proj1_from,
      y = proj2_from,
      xend = proj1_to,
      yend = proj2_to),
    color = "#000000") +
  geom_point(
    #size = 0.5,
    alpha = 0.2,
    color = '#a65628') +
  geom_segment(
    data=axes,
    aes(x=x1, y=y1, xend=x2, yend=y2),
    colour="grey70") +
  geom_text(
    data=axes,
    aes(x=x2, y=y2),
    label=rownames(axes),
    colour="grey50",
    size = 3) +
  geom_path(
    data=circle,
    aes(x=c1, y=c2), colour="grey70") +
  coord_fixed() +
  xlim(c(-0.025, 0.02)) +
  ylim(c(-0.025, 0.02)) +
  interior_annotation("a3", cex = 2)
```

```{r}
#| echo: false
#| eval: false
#| fig-cap: "Model in \\gD{}, on the layout a, and two views of the fit in projections from $50\\text{-}D$, for the PBMC data ($n =  5858$ and $p = 50$). Layout shows close clusters. In $50\\text{-}D$, the data shows "
#| label: fig-model-pbmc-author-proj-scdeed
#| fig-pos: H
#| fig-width: 15
#| fig-height: 5
#| out-width: 80%

free(trimesh_removed_pbmc) + 
  pbmc_proj_umap_model1 + 
  pbmc_proj_umap_model2 + 
  plot_layout(ncol=3) &
  theme(legend.position='none')
```
